---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
I load the data from activity.csv and convert the date collumn to class=date.
``` {r warning=FALSE, message=FALSE, cache=TRUE}

library(dplyr)
steps.df <- read.csv("activity.csv", stringsAsFactors = FALSE)
steps.df$date <-as.Date(steps.df$date)
```

## What is mean total number of steps taken per day?
``` {r warning=FALSE, message=FALSE, cache=TRUE}

stepsPerDay <- aggregate(steps ~  date, data=steps.df, sum)
hist(stepsPerDay$steps, breaks=100, main="Histogram of Steps per Day", xlab = "", ylim=range(0:5))
stepsPerDay.mean <- as.character(round(mean(stepsPerDay$steps)))
stepsPerDay.median <- as.character(median(stepsPerDay$steps))
```
The mean steps per day are `r stepsPerDay.mean` and the median is `r stepsPerDay.median`.

## What is the average daily activity pattern?
``` {r warning=FALSE, message=FALSE, cache=TRUE}

stepsPerDay <- aggregate(steps ~  date, data=steps.df, sum)
hist(stepsPerDay$steps, breaks=100, main="Histogram of Steps per Day", xlab = "", ylim=range(0:5))
stepsPerDay.mean <- mean(stepsPerDay$steps)
stepsPerDay.median <- median(stepsPerDay$steps)
stepsPerInterval <- aggregate(steps ~  interval, data=steps.df, mean)
plot(stepsPerInterval, type="l", main="Average Steps per Daily 5 Minute Time Interval", 
     at=seq(0,2400,200))
IntervalWithMostSteps <-stepsPerInterval[stepsPerInterval$steps==max(stepsPerInterval$steps),]
```
The interval with the most steps is `r IntervalWithMostSteps$interval` with `r round(IntervalWithMostSteps$steps,2)` steps.

## Imputing missing values
``` {r warning=FALSE, message=FALSE, cache=TRUE}
NACount <- sum(is.na(steps.df$steps))
```
There are `r NACount` records with missing/NA step values.
There are also many records with 0 values, but we do not consider those to be missing because it should mean that the person did not take any steps.

``` {r warning=FALSE, message=FALSE, cache=TRUE}
#Make a copy of data set and replace NA value with the average steps for that intval
NAfilled.df <- steps.df

for (i in 1:nrow(NAfilled.df)) {
  if (is.na(NAfilled.df[i,]$steps)) {
    meanSteps <- stepsPerInterval[stepsPerInterval$interval==NAfilled.df[i,]$interval,]
    NAfilled.df[i,]$steps <- meanSteps$steps
  }
}

stepsPerDay2 <- aggregate(steps ~  date, data=NAfilled.df, sum)
hist(stepsPerDay2$steps, breaks=100, main="Histogram of Steps per Day", xlab = "", ylim=range(0:5))
stepsPerDay2.mean <- as.character(round(mean(stepsPerDay2$steps)))
stepsPerDay2.median <- as.character(round(median(stepsPerDay2$steps)))

```
The mean steps per day for this data set are `r stepsPerDay2.mean` and the median is `r stepsPerDay2.median`. There is very little difference between the two datasets. The mean is the same,and in the second dataset the median is just one step higher and now equals the mean. I expected there to be more difference between the datasets so I wonder if I have made a mistake in my analysis. 

## Are there differences in activity patterns between weekdays and weekends?
``` {r warning=FALSE, message=FALSE, fig.height=6, cache=TRUE}
#Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
NAfilled.df$weekday <- ifelse((weekdays(NAfilled.df$date)=="Sunday") | 
          (weekdays(NAfilled.df$date)=="Saturday"), "Weekend", "Weekday")

stepsPerInterval.weekend <- aggregate(steps ~  interval, data=NAfilled.df[NAfilled.df$weekday=="Weekend",], mean)
stepsPerInterval.weekday <- aggregate(steps ~  interval, data=NAfilled.df[NAfilled.df$weekday=="Weekday",], mean)

par(mfrow = c(2, 1))

plot(stepsPerInterval.weekday, type="l", main="Weekday: Average Steps per Daily 5 Minute Time Interval", 
     axes = FALSE)
axis(side = 2, at = c(50,100,150,200))
axis(side = 1, at = seq(0,2400,200))

plot(stepsPerInterval.weekend, type="l", main="Weekend: Average Steps per Daily 5 Minute Time Interval", axes = FALSE)
axis(side = 2, at = c(50,100,150,200))
axis(side = 1, at = seq(0,2400,200))

```
I can see several differences between the trends on weekdays and weekends. On weekdays the steps begin around 5:30 AM, and on the weekends they do not reach that same level until around 8:00 am. 

Weekends and weekdays both spike around 8:15 am, though on weekends there are secondary spikes around noon and 4:00 PM. And on weekends activity is very low after 9:00 PM but on weekends there is a final skike around 10:00 PM.